The shape of kelp: environmental influences on kelp morphology

### 2018-09-26 chapter_1.Rmd
more added to wave exposure section
chnages and additions to chaper 1 and 2, also updated bibliography

Other than temperature, wave exposure is also recognised as an important driver of the marine environment, and macroalgae are not exception. Wave exposure has been shown to play a role in determing distribution, abundance, diversity, composition, growth [@Cousens1982] and productivity [@pedersen2012a] of macroalgae communties. For example, the width, vertical zonation and diversity of algal communities often change predictably along gradients of wave exposure. Wave exposure may also drive macroalgae communties indirectly through the alteration in affect of another environmental driver. For instance, increasing degrees of exposure may positively influence the amount of area available to trap light on macroalgal fronds, as well as increasing nutrient uptake due to increased turbulence in the boundry layer around the frond [@lobban1994]. The most important direct effect of wave exposure on macroalgal communities is through mechanical dislodgment, which ultimately leads to expiration. Wave exposure is a complex abiotic variable which varies spatially and temporarly in the marine enviroment. Furthermore, the degree to which a macroalgae community is exposed, is dependent on local site characteristics, such as bathymetry and local wind patterns. Despite this fact, macrolagae have been able to persist in often harsh and variable wave environments. Macroalage are sessile organisms and incapabable of migrating when local conditions become unsuitable. Therefore, macroalgae must adapt to the local wave climate in order to persist and survive, and achieve this through morphological adaptation. The morphology of macroalgae are not fixed gentic traits. A study by @?? showed that *species name* was able to rapidly adapt to the local wave environment when transplanted from a low wave exposure site to a high wave exposure site. Advances in genetic techniques and taxonomy have revealed that species delination based on morphology has been inaccurate, and organisms that were once considered two seperate species are actually one species.**insert example here**. The morphological adaption that macroalgae display are therefore driven by local or site conditions, and therefore individuals must be morphologically flexible to persist in stochastic environments. This may be achieved through different strategies and are species-specific, which can be directly attributed the high diversity in morphological characters of algal communities. For example some algae have fronds and others are articulated corraline, and therefore these species would need to adapt their morphology differently in order to persist. @moss1948 investigated the anatomy, chemical composistion of *Fucus spiralis* at three sites that varied in wave exposure (sheltered, medium exposure and exposed). The authors found that individuals in exposed sites showed less branching of thalli as well as variation in organic nitrogen, mannitol, laminarian and alginic acid concentrations. **insert more examples here**.

@@

Wave exposure may be modelled through various methods which range from simple cartographic to more advanced numerical wave models. Traditional ecological measures of wave exposure usually incorportates integrative measures of hyrdodynamic conditions at a particular site. More specifically, it is the integration of mechanical processes and the influence that is has on the ecology of nearshore communities. Cartographical models can be qualitative or quantitative and were designed for the need of wave exposure measures to explain ecological distributions. A simple set of calculations on coastline and wind data, and relatively small input data sets are required. These are regarded as "fetch-based models", which measure the length of open water associated with a site along a straight line. The output of such an approach is a simplified estimate of the potential wave energy for a specific set of sites. Advances in cartographical methods using fetch-based models has allowed for wave exposure measurments for larger areas, and has been suggested as a method for predicting macroalgal community structure [@burrows2008]. An example of such a model is the "BioEx model" which was developed by @baardseth1970 to estimate wave exposure over large regions. BioEx requires frequency, strength and direction of winds, weighted by degree of exposure within various directions. BioEx is calculated as the sum of the index developed at different spatial scales (local, fjord and open). This method has been used in mapping of marine coastal biodiveristy [@Rinde2004].

@lindegarth2005 critized this approach, arguing that the choice of wave exposure method  can influence ecological inference. The authors also highlighted the need for objective, reproducible and quantitative studies comparing exposure indices [@lindegarth2005]. A study by @sundbald...**include description of study here**...Other authors, such as @hill2010, have argued that these simple measures can be improved upon by including bathymetery data which allows the incorporation of diffraction into the calculation. Diffraction is topographically induced variations in wave direction. A model incorporating this complexity was developed by @isaeus2004, and is known as the "simplified wave model" (SWM). The model uses measurements of wind strength, fetch and empirically derived algorithims to mimic diffraction.

Advances in numerical modelling have been founded on physical wave theory on how a wave "behaves". This approach is based on a theoretical persepctive rather than the need to answer ecological questions. Besides diffraction, these models incorporate more complexity by including wind forcing, wave-to-wave interactions and loss of energy due to friction and wave breaking. Numerical models have a variety of applications and are often incorporated within hydrodynamic general circulation models and are used operationally for forecasting the sea state [@hasselmann1988; @booij1999; @smith2001]. The downside of advanced numerical models is that are computationally intensive which creates limitations for large scale simulations. Therefore, their application along long stretches of variable coastline, inshore environments and ocean-wide simulations limited due to the poor spatial coverage. However, this models can be designed for local or site specific coverage, provided the correct data is available.

>>>>

Other than temperature, wave exposure is also recognised as an important driver of the marine environment, and macroalgae are not exception. Wave exposure has been shown to play a role in determing distribution, abundance, diversity, composition, growth [@Cousens1982] and productivity [@pedersen2012a] of macroalgae communties. For example, the width, vertical zonation and diversity of algal communities often change predictably along gradients of wave exposure. Wave exposure may also drive macroalgae communties indirectly through the alteration in affect of another environmental driver. For instance, increasing degrees of exposure may positively influence the amount of area available to trap light on macroalgal fronds, as well as increasing nutrient uptake due to increased turbulence in the boundry layer around the frond [@lobban1994]. The most important direct effect of wave exposure on macroalgal communities is through mechanical dislodgment, which ultimately leads to expiration. Wave exposure is a complex abiotic variable which varies spatially and temporarly in the marine enviroment. Furthermore, the degree to which a macroalgae community is exposed, is dependent on local site characteristics, such as bathymetry and local wind patterns. Despite this fact, macrolagae have been able to persist in often harsh and variable wave environments. Macroalage are sessile organisms and incapabable of migrating when local conditions become unsuitable. Therefore, macroalgae must adapt to the local wave climate in order to persist and survive, and achieve this through morphological adaptation. The morphology of macroalgae are not fixed gentic traits. A study by @?? showed that *species name* was able to rapidly adapt to the local wave environment when transplanted from a low wave exposure site to a high wave exposure site. Advances in genetic techniques and taxonomy have revealed that species delination based on morphology has been inaccurate, and organisms that were once considered two seperate species are actually one species.**insert example here**. The morphological adaption that macroalgae display are therefore driven by site conditions, therefore individuals must be morphologically flexible to persist in stochastic environments. This may be achieved through different strategies and are species-specific, which can be directly attributed the high diversity in morphological characters of algal communities. For example some algae have fronds and others are articulated corraline, and therefore these species would need to adapt their morphology differently in order to persist. In general, flat strap-like blades are common in areas that are exposed to high wave energy, while at proected sites blade morphology is wide and undulated.

@moss1948 investigated the anatomy, chemical composistion of *Fucus spiralis* at three sites that varied in wave exposure (sheltered, medium exposure and exposed). The authors found that individuals in exposed sites showed less branching of thalli as well as variation in organic nitrogen, mannitol, laminarian and alginic acid concentrations.

@@

Wave exposure may be modelled through various methods which range from simple cartographic to more advanced numerical wave models. Traditional ecological measures of wave exposure usually incorportates integrative measures of hyrdodynamic conditions at a particular site. More specifically, it is the integration of mechanical processes and the influence that is has on the ecology of nearshore communities. Cartographical models can be qualitative or quantitative and were designed for the need of wave exposure measures to explain ecological distributions. A simple set of calculations on coastline and wind data, and relatively small input data sets are required. These are regarded as "fetch-based models", which measure the length of open water associated with a site along a straight line. The output of such an approach is a simplified estimate of the potential wave energy for a specific set of sites. Advances in cartographical methods using fetch-based models has allowed for wave exposure measurments for larger areas, and has been suggested as a method for predicting macroalgal community structure [@burrows2008]. An example of such a model is the "BioEx model" which was developed by @baardseth1970 to estimate wave exposure over large regions. BioEx requires frequency, strength and direction of winds, weighted by degree of exposure within various directions. BioEx is calculated as the sum of the index developed at different spatial scales (local, fjord and open). This method has been used in mapping of marine coastal biodiveristy [@Rinde2004]. @lindegarth2005 critized this approach, arguing that the choice of wave exposure method  can influence ecological inference. The authors also highlighted the need for objective, reproducible and quantitative studies comparing exposure indices [@lindegarth2005]. A study by @sundbald...**include description of study here**...Other authors, such as @hill2010, have argued that these simple measures can be improved upon by including bathymetery data which allows the incorporation of diffraction into the calculation. Diffraction is topographically induced variations in wave direction. A model incorporating this complexity was developed by @isaeus2004, and is known as the "simplified wave model" (SWM). The model uses measurements of wind strength, fetch and empirically derived algorithims to mimic diffraction.

Advances in numerical modelling have been founded on physical wave theory on how a wave "behaves". This approach is based on a theoretical perspective rather than the need to answer ecological questions. Besides diffraction, these models incorporate more complexity by including wind forcing, wave-to-wave interactions and loss of energy due to friction and wave breaking. Numerical models have a variety of applications and are often incorporated within hydrodynamic general circulation models and are used operationally for forecasting the sea state [@hasselmann1988; @booij1999; @smith2001]. The downside of advanced numerical models is that are computationally intensive which creates limitations for large scale simulations. Therefore, their application along long stretches of variable coastline, inshore environments and ocean-wide simulations limited due to the poor spatial coverage. However, this models can be designed for local or site specific coverage, provided the correct data is available.

### 2018-09-26 chapter_1.tex
chnages and additions to chaper 1 and 2, also updated bibliography

Other than temperature, wave exposure is also recognised as an important
driver of the marine environment, and macroalgae are not exception. Wave
exposure has been shown to play a role in determing distribution,
abundance, diversity, composition, growth {[}@Cousens1982{]} and
productivity {[}@pedersen2012a{]} of macroalgae communties. For example,
the width, vertical zonation and diversity of algal communities often
change predictably along gradients of wave exposure. Wave exposure may
also drive macroalgae communties indirectly through the alteration in
affect of another environmental driver. For instance, increasing degrees
of exposure may positively influence the amount of area available to
trap light on macroalgal fronds, as well as increasing nutrient uptake
due to increased turbulence in the boundry layer around the frond
{[}@lobban1994{]}. The most important direct effect of wave exposure on
macroalgal communities is through mechanical dislodgment, which
ultimately leads to expiration. Wave exposure is a complex abiotic
variable which varies spatially and temporarly in the marine enviroment.
Furthermore, the degree to which a macroalgae community is exposed which
depend on local site characteristics, such as bathymetry and local wind
patterns. Despite this fact, macrolagae have been able to persist in
often harsh and variable wave environments. Macroalage are sessile
organisms and incapabable of migrating when local conditions become
unsuitable. Therefore, macroalgae must adapt to the local wave climate
in order to persist and survive, and achieve this through morphological
variation.

\begin{quote}
keeping this for now @moss1948 investigated the anatomy, chemical
composistion of \emph{Fucus spiralis} at three sites that varied in wave
exposure (sheltered, medium exposure and exposed). The authors found
that individuals in exposed sites showed less branching of thalli as
well as variation in organic nitrogen, mannitol, laminarian and alginic
acid concentrations.
\end{quote}

@@

Traditional ecological measures of wave exposure usually incorportates
integrative measures of hyrdodynamic conditions at a particular site.
More specifically, it is the integration of mechanical processes and the
influence that is has on the ecology of nearshore communities.

Wave exposure may be modelled through various methods which range from
simple cartographic to more advanced numerical wave models.
Cartographical models can be qualitative or quantitative and were
designed for the need of wave exposure measures to explain ecological
distributions. A simple set of calculations on coastline and wind data,
and relatively small input data sets are required. These are regarded as
``fetch-based models'', which measure the length of open water
associated with a site along a straight line. The output of such an
approach is a simplified estimate of the potential wave energy for a
specific set of sites. Advances in cartographical methods using
fetch-based models has allowed for wave exposure measurments for larger
areas, and has been suggested as a method for predicting macroalgal
community structure {[}@burrows2008{]}. An example of such a model is
the ``BioEx model'' which was developed by @baardseth1970 to estimate
wave exposure over large regions. BioEx requires frequency, strength and
direction of winds, weighted by degree of exposure within various
directions. BioEx is calculated as the sum of the index developed at
different spatial scales (local, fjord and open). This method has been
used in mapping of marine coastal biodiveristy {[}@Rinde2004{]}.

>>>>

Other than temperature, wave exposure is also recognised as an important
driver of the marine environment, and macroalgae are not exception. Wave
exposure has been shown to play a role in determing distribution,
abundance, diversity, composition, growth {[}@Cousens1982{]} and
productivity {[}@pedersen2012a{]} of macroalgae communties. For example,
the width, vertical zonation and diversity of algal communities often
change predictably along gradients of wave exposure. Wave exposure may
also drive macroalgae communties indirectly through the alteration in
affect of another environmental driver. For instance, increasing degrees
of exposure may positively influence the amount of area available to
trap light on macroalgal fronds, as well as increasing nutrient uptake
due to increased turbulence in the boundry layer around the frond
{[}@lobban1994{]}. The most important direct effect of wave exposure on
macroalgal communities is through mechanical dislodgment, which
ultimately leads to expiration. Wave exposure is a complex abiotic
variable which varies spatially and temporarly in the marine enviroment.
Furthermore, the degree to which a macroalgae community is exposed, is
dependent on local site characteristics, such as bathymetry and local
wind patterns. Despite this fact, macrolagae have been able to persist
in often harsh and variable wave environments. Macroalage are sessile
organisms and incapabable of migrating when local conditions become
unsuitable. Therefore, macroalgae must adapt to the local wave climate
in order to persist and survive, and achieve this through morphological
adaptation. The morphology of macroalgae are not fixed gentic traits. A
study by @?? showed that \emph{species name} was able to rapidly adapt
to the local wave environment when transplanted from a low wave exposure
site to a high wave exposure site. Advances in genetic techniques and
taxonomy have revealed that species delination based on morphology has
been inaccurate, and organisms that were once considered two seperate
species are actually one species.\textbf{insert example here}. The
morphological adaption that macroalgae display are therefore driven by
site conditions, therefore individuals must be morphologically flexible
to persist in stochastic environments. This may be achieved through
different strategies and are species-specific, which can be directly
attributed the high diversity in morphological characters of algal
communities. For example some algae have fronds and others are
articulated corraline, and therefore these species would need to adapt
their morphology differently in order to persist. In general, flat
strap-like blades are common in areas that are exposed to high wave
energy, while at proected sites blade morphology is wide and undulated.

@moss1948 investigated the anatomy, chemical composistion of \emph{Fucus
spiralis} at three sites that varied in wave exposure (sheltered, medium
exposure and exposed). The authors found that individuals in exposed
sites showed less branching of thalli as well as variation in organic
nitrogen, mannitol, laminarian and alginic acid concentrations.

@@

Wave exposure may be modelled through various methods which range from
simple cartographic to more advanced numerical wave models. Traditional
ecological measures of wave exposure usually incorportates integrative
measures of hyrdodynamic conditions at a particular site. More
specifically, it is the integration of mechanical processes and the
influence that is has on the ecology of nearshore communities.
Cartographical models can be qualitative or quantitative and were
designed for the need of wave exposure measures to explain ecological
distributions. A simple set of calculations on coastline and wind data,
and relatively small input data sets are required. These are regarded as
``fetch-based models'', which measure the length of open water
associated with a site along a straight line. The output of such an
approach is a simplified estimate of the potential wave energy for a
specific set of sites. Advances in cartographical methods using
fetch-based models has allowed for wave exposure measurments for larger
areas, and has been suggested as a method for predicting macroalgal
community structure {[}@burrows2008{]}. An example of such a model is
the ``BioEx model'' which was developed by @baardseth1970 to estimate
wave exposure over large regions. BioEx requires frequency, strength and
direction of winds, weighted by degree of exposure within various
directions. BioEx is calculated as the sum of the index developed at
different spatial scales (local, fjord and open). This method has been
used in mapping of marine coastal biodiveristy {[}@Rinde2004{]}.

### 2018-09-26 chapter_2.Rmd
chnages and additions to chaper 1 and 2, also updated bibliography

Kelps are sessile species and have to adapt to local environmental conditions in order to persist and survive. The resilence of kelps is evident through their distribution, which spans a range of environmental conditions that vary spatially and temporarly. Wave exposure and temperature are regarded as important environmental drivers of kelps, which play a role in the growth, reproduction, and survival of kelp species. Wave exposure is recognised as the main cause of kelp mortality, which may manifest itself through dislodgment of individuals or breakage of important morphological structures. These mechanisms usually occur during times of high wave energy or pulse disturbance events, such as storms. Kelps are able to reduce the probablity of dislodgment or breakage through morphological adaptation, which may involve several strategies. Kelps can either reduce drag, increase attachment strength or increase flexibility in order to survive high wave energy environments. **insert examples??*

>>>>

Kelps are sessile species and have to adapt to local environmental conditions in order to persist and survive. The resilence of kelps is evident through their distribution, which spans a range of environmental conditions that vary spatially and temporarly. Wave exposure and temperature are regarded as important environmental drivers with regards to macroalgae, and play a role in the growth, reproduction, and survival of kelp species. Wave exposure is recognised as the main cause of kelp mortality, which may manifest itself through dislodgment of individuals or breakage of important morphological structures. These mechanisms usually occur during times of high wave energy or pulse disturbance events, such as storms. Kelps are able to reduce the probablity of dislodgment or breakage through morphological adaptation, which may involve several strategies. Kelps can either reduce drag, increase attachment strength or increase flexibility in order to survive high wave energy environments.
